Supplementary Materials

### 2023-06-30 manuscript/manuscript.tex
methods

\begin{align}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \nonumber \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)}.
\end{align}

@@

where $V_a = (U-R)s$ as in Equation \eqref{eq:Va}. The predicted $V_a$, $R$,
and $N_d$ values represent the equilibria expected under a
mutation-selection-drift-draft process for neutral to strongly deleterious
additive mutations. The collection of these $V_a$ and $R$ equilibria can then
be used to calculate the reduction at an arbitrary position in the genome (see
Methods

@@

We use these pre-computed equilibria to calculate the reduction map $B(x; m_i,
s_j)$ at positions $x$ across the mutation grids described above. Since we
assume multiplicative fitness, the reduction is the product of each segment's
contribution accounting for the recombination is,

\begin{align}
    B(x; m_i, s_j) = \exp\left(- \frac{1}{2}\sum_{g \in G} \sum_{i=1}^{n_s} V_g(m_i, s_j) Q_g^2(m_i, s_j, r_{x, g})\right)
\end{align}

where $r_{x, g}$ is the recombination fraction between the focal site and
segment $g$. A separate reduction map is calculated for all features $G$ within
a specific feature type. Our B' maps are calculated for $\log_{10}$-spaced
grids over $10^{-1} \le s \le 10^{-8}$ and $10^{-11} \le m \le 10^{-7}$, in
10kb increments across the genome.

\subsection*{Composite Likelihood}
\label{sec:methods-likelihood}

\begin{align}
    \log\left(B(x; W, s_j) \right) = - \frac{1}{2} \sum_{g \in G} \sum_{i=1}^{n_s} V_g(\mu W_{j, k(g)}, s_j) Q_g^2(m_i, s_j, r_{x, g})
\end{align}

>>>>

\begin{align}
    \label{eq:Q}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \nonumber \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)}.
\end{align}

@@

where $V_a = (U-R)s$ as in Equation \eqref{eq:Va}. We denote the solutions to
these equation, which represent equilibria under mutation-selection-drift-draft
process, as $\widetilde{N}_d$ and $\widetilde{R}$. These equilibria also imply
an equilibrium level of additive fitness variation $\widetilde{V}_a$ in the
segment, are used to calculate the reduction at an arbitrary position in the
genome (see Methods

@@

Our method uses the pre-computed equilibria $\widetilde{V}_{a,g}$ for each
segment $g$ to calculate the reduction map $B(x; m_i, s_j)$ at positions $x$
across the parameter grids described above. Since we assume multiplicative
fitness, the reduction is the product of each segment's contribution accounting
for the recombination is,

\begin{align}
    B(x; m_i, s_j) = \exp\left(- \frac{1}{2}\sum_{g \in G} \sum_{i=1}^{n_s} \widetilde{V}_{a,g}(m_i, s_j) Q_g^2(m_i, s_j, r_{x, g})\right)
\end{align}
%
where $r_{x, g}$ is the recombination fraction between the focal site and
segment $g$. Here, $Q^2(m_i, s_j, r_{x,g})$ is given by Equation \eqref{eq:Q}
squared. A separate reduction map is calculated for all features $G$ within a
specific feature type. We calculate B' calculate for $\log_{10}$-spaced grids
over $10^{-1} \le s \le 10^{-8}$ and $10^{-11} \le m \le 10^{-7}$, in 10kb
increments across the genome.

\subsection*{Composite Likelihood and Optimization}
\label{sec:methods-likelihood}

Following previous approaches
\parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}, we use a composite
likelihood approach to fit our negative selection model. Per-basepair allele
count data (described below) is summarized into the number of same and
different pairwise differences per window. All of our primary models were fit
with megabase windows, since previous work has found the strongest selection
signal at this scale (we confirm this with one XXX fit at the $10^5$bp scale).

We use a binomial likelihood with probability $\bar{\pi}(b; \Psi) = \pi_0
\bar{B}(b; \mu, \mathbf{W})$, where bars indicate averages over some bin width.
With multiple features classes and a map $k(g)$ between a segment $g$ to its
class, the reduction map is,

\begin{align}
    \log\left(B(x; W, s_j) \right) = - \frac{1}{2} \sum_{g \in G} \sum_{i=1}^{n_s} V_g(\mu W_{j, k(g)}, s_j) Q_g^2(m_i, s_j, r_{x, g}).
\end{align}

Here, $\mu$ is the mutation rate and $\mathbf{W}$ is the $n_s \times K$ DFE
matrix over the selection grid and features. Since each column of the DFE
$\mathbf{W}$ must sum to one, numeric optimization was done in unconstrained
$n_s-1$ dimensional softmax space. Our binomial likelihood modeled the number
of different pairwise comparisons observed per window given the predicted
pairwise diversity and total number of comparisons implied per window (see
Supplementary Materials Section \ref{supp:data-summary}).

We tested multiple different optimization routines, finding that BOBYQA
outperformed alternatives \parencite{Powell_undated-ec}. We inspected and
confirmed convergence with diagnostic plots finding stable optima across 10,000
random starts (see Supplementary Materials Section XXX).

### 2023-06-30 manuscript/supplementary.tex
methods

\begin{align}
  \mathbf{M}_\text{S} = \mu \begin{bmatrix}
    w_{10^{-6},\text{CDS}} & w_{10^{-6},\text{UTR}} & w_{10^{-6},\text{PC}} \\
    w_{10^{-5},\text{CDS}} & w_{10^{-5},\text{UTR}} & w_{10^{-5},\text{PC}} \\
    \vdots & \vdots & \vdots \\
    w_{10^{-1},\text{CDS}} & w_{10^{-1},\text{UTR}} & w_{10^{-1},\text{PC}} \\
  \end{bmatrix}
\end{align}
%
which has $2 + K \times (n_s-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization. We implement this using softmax, which describe 
further in Supplementary Section \ref{supp:optim}.

@@

\subsection{Data Summary Matrices}

@@

These neutral tracks are all produced by the Snakemake file
\path{data/fit_annotation/Snakefile} in this study's GitHub repository
(\url{http://github.com/vsbuffalo/bprime}).

\begin{table}
\centering
\caption{Percent of basepairs remaining that are accessible and 
         putatively neutral.}
\begin{tabular}{|l|rrr|}
\hline
Chromosome & Accessible & Neutral & Intersection \\
\hline
 chr1 &        39.7 &     61.2 &  23.4 \\
 chr2 &        45.3 &     67.3 &  27.5 \\
 chr3 &        43.2 &     67.3 &  25.7 \\
 chr4 &        42.7 &     70.5 &  27.4 \\
 chr5 &        42.4 &     67.5 &  26.0 \\
 chr6 &        43.8 &     67.9 &  26.7 \\
 chr7 &        41.9 &     67.8 &  26.2 \\
 chr8 &        42.3 &     70.9 &  27.5 \\
 chr9 &        36.2 &     58.1 &  22.0 \\
chr10 &        42.9 &     68.2 &  26.8 \\
chr11 &        41.3 &     67.0 &  25.0 \\
chr12 &        40.6 &     67.2 &  24.1 \\
chr13 &        38.8 &     59.8 &  25.2 \\
chr14 &        36.5 &     54.9 &  21.5 \\
chr15 &        34.4 &     53.7 &  20.6 \\
chr16 &        36.0 &     62.5 &  23.1 \\
chr17 &        37.2 &     63.3 &  21.6 \\
chr18 &        40.3 &     65.7 &  26.3 \\
chr19 &        28.1 &     66.1 &  17.0 \\
chr20 &        36.1 &     65.1 &  22.4 \\
chr21 &        29.8 &     55.9 &  20.9 \\
chr22 &        25.5 &     50.9 &  17.1 \\
\hline
\end{tabular}
\end{table}

@@

$\Phi$ is the set of BGS parameters (i.e. the DFE for each feature type and
mutation rate), and $\pi_0 = 4 N_e \mu$ is determined by the genome-wide
drift-effective population size $N_e$, set by only reproductive and demographic
processes. Regional mutation rate heterogeneity can be accounted for with a
regional mutation rate scaling function $m(x)$, $\pi(x) = B(x, \Phi) m(x)
\pi_0$, but we do not explore mutation-rate heterogeneity, as it is unclear how
to differentiate variance in mutation rate from the substitution rate
heterogeneity we find across the genome.

@@

>>>>

\begin{align}
    \mathbf{M} = \mu \mathbf{W} &= \begin{bmatrix} 
    w_{10^{-6},\text{CDS}} & w_{10^{-6},\text{UTR}} & w_{10^{-6},\text{PC}} \\
    w_{10^{-5},\text{CDS}} & w_{10^{-5},\text{UTR}} & w_{10^{-5},\text{PC}} \\
    \vdots & \vdots & \vdots \\
    w_{10^{-1},\text{CDS}} & w_{10^{-1},\text{UTR}} & w_{10^{-1},\text{PC}} \\
  \end{bmatrix}
\end{align}
%
which has $2 + K \times (n_s-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization. We implement this using softmax, which describe 
further in Supplementary Section \ref{supp:optim}.

@@

\subsection{Data Summary Matrices}
\label{supp:data-summary}

@@

These neutral tracks are all produced by the Snakemake file
\path{data/fit_annotation/Snakefile} in this study's GitHub repository
(\url{http://github.com/vsbuffalo/bprime}). We summarize the percentage of
basepairs after masking per chromosome in Supplementary Table
\ref{supp-table:masks}.

\begin{table}
    \label{supp-table:masks}
\centering
\caption{Accessible and putatively neutral sequence mask statistics, for the criteria used in 
         the main model fits.}
\begin{tabular}{|c|cccc|}
\hline
Chromosome & Accessible (\%) & Neutral (\%) & Intersection (\%) & Total Basepairs ($\times 10^6$) \\
\hline 
1 &        39.7 &     62.0 &  15.7 &     39.09 \\
2 &        45.3 &     60.1 &  18.9 &     45.77 \\
3 &        43.2 &     59.6 &  16.9 &     33.51 \\
4 &        42.7 &     65.9 &  20.2 &     38.42 \\
5 &        42.4 &     62.2 &  18.1 &     32.86 \\
6 &        43.8 &     61.4 &  18.4 &     31.43 \\
7 &        41.9 &     64.0 &  18.6 &     29.64 \\
8 &        42.3 &     66.1 &  20.2 &     29.32 \\
9 &        36.2 &     65.4 &  14.6 &     20.21 \\
10 &        42.9 &     62.2 &  18.8 &     25.15 \\
11 &        41.3 &     61.5 &  17.0 &     22.96 \\
12 &        40.6 &     60.4 &  15.9 &     21.19 \\
13 &        38.8 &     70.7 &  18.8 &     21.50 \\
14 &        36.5 &     64.6 &  13.8 &     14.77 \\
15 &        34.4 &     66.3 &  13.5 &     13.77 \\
16 &        36.0 &     66.0 &  15.5 &     14.00 \\
17 &        37.2 &     58.8 &  14.2 &     11.82 \\
18 &        40.3 &     66.1 &  19.6 &     15.75 \\
19 &        28.1 &     67.5 &  11.5 &      6.74 \\
20 &        36.1 &     62.5 &  15.4 &      9.92 \\
21 &        29.8 &     76.9 &  17.3 &      8.08 \\
22 &        25.5 &     75.5 &  12.6 &      6.40 \\
\hline
weighted average & 40.1 & 63.6 & 17.1 & - \\
\hline
\end{tabular}
\end{table}

@@

$\Phi$ is the set of BGS parameters (i.e. the DFE matrix $\mathbf{W}$ and
mutation rate $\mu$), and $\pi_0 = 4 N_e \mu$ is determined by the genome-wide
drift-effective population size $N_e$, set by only reproductive and demographic
processes. Regional mutation rate heterogeneity can be accounted for with a
regional mutation rate scaling function $m(x)$, $\pi(x) = B(x, \Phi) m(x)
\pi_0$, but we do not explore mutation-rate heterogeneity, as it is unclear how
to differentiate variance in mutation rate from the substitution rate
heterogeneity we find across the genome.

@@

\subsection{Optimization Diagnostics}

\begin{figure}[!htb]
  \label{suppfig:diag-plot}
  \centering
  \includegraphics[width=\textwidth]{figures/supplementary/figure_feature_priority_yri_full_diag.pdf}
  \caption{
      An optimization diagnostic plot showing the }
\end{figure}
